---
title: "NGS Oncology: impact of low DNA input and tumor purity"
author: "Eugenio Graceffo"
date: "June-July 2024"
output:
  html_document:
    toc: True
    toc_float: True
    df_print: paged
---
## 
**Project 1**: what is the impact of low amount of DNA as starting material? \
**Project 2**: what is the impact of tumor purity on the detected mutations?



```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE, results='hide'}
#load libraries
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(matrixStats)
library(ggrepel)
library(scales)
library(ggpubr)
library(readxl)

library(xml2)
library(XML)


library(data.table)

library(pheatmap)

```



# Data input
## PASS mutations count from .vcf

First, lets extract the number of detected PASS mutations from the .vcf files
```{r}
## set the directory where the vcf file are
path_to_sequencing_folder <- "C:/Users/user1/Desktop/OncoPanel_project/Sequencing_Data"


# initialize an empty dataframe for PASS counts
vcf_mutation_counts <- data_frame("Sample"=character(),
                   "num_mutations"=double()) 

# initialize an empty data frame for the Mutations ID dataframe
vcf_mutation_ID <- data_frame("mutation_ID"=character()) 


for (x in list.files(path_to_sequencing_folder)) {
  
  muton_path <- paste0(path_to_sequencing_folder,"/", x, "/REPORT_", x, "/muton.vcf")
  table_sample <- fread(muton_path) ## vcf file
  table_sample_filtered <- filter(table_sample, table_sample$FILTER == "PASS") ## get only positive mutations
  
  ## mutation COUNTS df
  new_row <- c(x, nrow(table_sample_filtered)) ## generate a new row to add to the output dataframe
  vcf_mutation_counts <- rbind(vcf_mutation_counts, new_row)  ## append to dataframe
  
  ## mutation ID df
  sample_name <- as.character(x)
  table_sample_filtered_ID <- table_sample_filtered %>%
    unite("mutation_ID", c(1,2,4,5), remove = FALSE) %>% ## generates a column with a unique ID specific to the mutation
    select(mutation_ID)
  
  table_sample_filtered_ID[[sample_name]] <- TRUE
  
  vcf_mutation_ID <- full_join(vcf_mutation_ID, table_sample_filtered_ID, by= "mutation_ID", )  ## append to dataframe
}




## rename column names
vcf_mutation_counts <-rename(vcf_mutation_counts, Sample = 1)
vcf_mutation_counts <- rename(vcf_mutation_counts, num_mutations = 2)

## adjust values type
vcf_mutation_counts$num_mutations <- as.double(vcf_mutation_counts$num_mutations)

vcf_mutation_counts


## code for individual loading
#muton_vcf <- fread("C:/Users/user1/Desktop/T03_original_data/REPORT_03/fusion.vcf")
#muton_vcf

#filtered_vcf <- muton_vcf %>%
  #filter(muton_vcf$FILTER == "PASS")

#filtered_vcf

vcf_mutation_ID


```



## Total counts from report.xml
```{r}
# initialize an empty dataframe
QC_total_counts <- data_frame() 


for (x in list.files(path_to_sequencing_folder)) {
  
  QC_path <- paste0(path_to_sequencing_folder,"/", x, "/report.xml")
  QC_data <- read_xml(QC_path)
  QC_xml <- xmlParse(QC_data)
  df_QC <- xmlToDataFrame(nodes = getNodeSet(QC_xml, "//report/qc-results/tumor"))  ## extract the node containing the QC info
  QC_total_counts <- rbind(QC_total_counts,df_QC)  ##append to dataframe

  
}

## adjust the values type
QC_total_counts$total_reads <- as.double(gsub(",", "", QC_total_counts$total_reads))
QC_total_counts$`mapping-rate` <- as.double(QC_total_counts$`mapping-rate`)
QC_total_counts$Duplication_rate <- as.double(QC_total_counts$`duplication-rate`)
QC_total_counts$total_reads <- as.double(QC_total_counts$total_reads)

QC_total_counts <- QC_total_counts %>%
  mutate(reads_name = paste0(QC_total_counts$reads_name, "_N03")) %>% ## add the missing "_N03" to match sample names in other dataframes
  rename(Sample = reads_name) %>%
  select(-panel, -reagent, -run_date, -`duplication-rate`)



QC_total_counts


```


## Counts used for pipeline from validation.xml
```{r}
# initialize an empty dataframe
Unique_counts <- data_frame() 


for (x in list.files(path_to_sequencing_folder)) {
  
  Validation_path <- paste0(path_to_sequencing_folder,"/", x, "/validation.xml")
  Validation_data <- read_xml(Validation_path)
  Validation_xml <- xmlParse(Validation_data)
  df_validation <- xmlToDataFrame(nodes = getNodeSet(Validation_xml, "//report/metrics_target_regions/reads"))  ## extract the node containing the QC info
  new_row_validation <- c(x, df_validation$text) ## generate a new row to add to the output dataframe
  Unique_counts <- rbind(Unique_counts,new_row_validation)  ##append to dataframe

  
}

## rename column names
Unique_counts <-rename(Unique_counts, Sample = 1)
Unique_counts <- rename(Unique_counts, unique_reads = 2)


## adjust the values type
Unique_counts$unique_reads <- as.double(gsub(",", "", Unique_counts$unique_reads))
Unique_counts$unique_reads <- as.double(Unique_counts$unique_reads)

Unique_counts


```


## Metadata info .xlsx
```{r}
# load the metadata info
metadata <- read_excel("metadata.xlsx")
metadata
```

```{r}
df <- metadata%>%
  inner_join(vcf_mutation_counts) %>%
  inner_join(QC_total_counts) %>%
  inner_join(Unique_counts)

df
```

## AF Extrapolation from muton.vcf
### Dataframes containing AF values 

First, lets load up all the .vcf files into a list of dataframes
```{r}

# initialize an empty list of dataframes
dataframe_list <- list()

for (x in list.files(path_to_sequencing_folder)) {
  
  muton_path <- paste0(path_to_sequencing_folder,"/", x, "/REPORT_", x, "/muton.vcf")
  table_sample <- fread(muton_path) ## vcf file
  table_sample$Sample <- x  ## creates a column with the sample name
  table_sample <- select(table_sample, -(str_remove(x, "_N03")))
  new_row <- c(x, nrow(table_sample_filtered)) ## generate a new row to add to the output dataframe

  dataframe_list[[x]] <- table_sample  ## add dataframe to the list
  
}

```


### AF Extrapolation function
Lets write a function to expand the INFO tab containing the AF values into more columns
```{r}
## function to expand INFO tab
expand_info_tab <- function(vcf_file){
  vcf_file %>%
  unite("mutation_ID", c(1,2,4,5), remove = FALSE) %>% ## generates a column with a unique ID specific to the mutation
  separate_rows(INFO, sep = ";") %>% ## extracts info from cell and spreads it over dataframe
  separate(INFO, c('col1', 'col2'), sep = '=') %>% ## separates the key and the value of the dictionary
  pivot_wider(names_from = col1, values_from = col2) ## puts the key as column names on the original dataframe

}

```

### For loop for extrapolation and function application
And use the function to extract the AF values for all the dataframes
```{r}

df_AF_expanded <- data_frame() 

for (x in list.files(path_to_sequencing_folder)) {
  muton_vcf <- dataframe_list[[x]]
  muton_vcf_expanded <- expand_info_tab(muton_vcf) 
  
  #＃adjust column type
  muton_vcf_expanded$AF <- as.double(muton_vcf_expanded$AF)
  muton_vcf_expanded$FPV <- as.double(muton_vcf_expanded$FPV)
  muton_vcf_expanded$FQV <- as.double(muton_vcf_expanded$FQV)
  muton_vcf_expanded$DP <- as.double(muton_vcf_expanded$DP)
  

  df_AF_expanded <- rbind(df_AF_expanded, muton_vcf_expanded)  ## append to dataframe
  
}

df_AF_expanded


```

### Adding metadata information
```{r}
df_AF_total <- metadata%>%
  inner_join(df_AF_expanded) %>%
  select(-Total_DNA) %>%
  separate(Sample_name, into = "Total_DNA", extra = "drop",  remove = FALSE) 

df_AF_final <- df_AF_total %>%
  filter(Ratio_experiment == "yes")

## adjust types for graphing
df_AF_final$Total_DNA <- factor(df_AF_final$Total_DNA, levels = c("200ng", "50ng", "30ng"))
df_AF_final$Tumor_percentage <- factor(df_AF_final$Tumor_percentage, levels = c("100_%", "50_%", "25_%"))



df_AF_final
```



# Project 1: Impact of lower amounts of DNA

```{r}
DNA_experiment <- df %>%
  filter(Tumor_percentage == "100_%" ) %>%
  arrange(desc(Total_DNA))



DNA_experiment

```

## Total reads

```{r}

## graph the combined results
ggscatter(DNA_experiment, x= "Total_DNA", y="total_reads",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "gray", fill = "lightgray", linetype= "solid"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 47500000),
          cor.coef.size = 5,
          title = "Impact on total reads",
          
          ) +
  ylab("Total number of reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = 45000000, linetype = "dashed", color = "red", linewidth= 0.7) +
  geom_text(aes(181, 45000000, label = "Threshold 45M", vjust = -1), color = "red")



## graph the individual results
ggscatter(DNA_experiment, x= "Total_DNA", y="total_reads",
          color = "black", shape = , size = 3, facet.by = "Experimenter",
          add = "reg.line",
          add.params = list(color = "gray", fill = "lightgray", linetype= "solid"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 48500000),
          cor.coef.size = 5,
          title = "Impact on total reads",
          
          ) +
  ylab("Total number of reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = 45000000, linetype = "dashed", color = "red", linewidth= 0.7) +
  geom_text(aes(181, 45000000, label = "Threshold 45M", vjust = -1), color = "red")
  
```

## Unique reads

```{r}

## graph the combined results
ggscatter(DNA_experiment, x= "Total_DNA", y="unique_reads",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "gray", fill = "lightgray", linetype= "solid"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 30500000),
          cor.coef.size = 5,
          title = "Impact on unique reads used for OncoPanel pipeline",
          
          ) +
  ylab("Total number of unique reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = 45000000, linetype = "dashed", color = "red", linewidth= 0.7) +
  geom_text(aes(170, 40000000, label = "Threshold 45M", vjust = -1), color = "red")



## graph the individual results
ggscatter(DNA_experiment, x= "Total_DNA", y="unique_reads",
          color = "black", shape = , size = 3, facet.by = "Experimenter",
          add = "reg.line",
          add.params = list(color = "gray", fill = "lightgray", linetype= "solid"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 30500000),
          cor.coef.size = 5,
          title = "Impact on unique reads used for OncoPanel pipeline",
          
          ) +
  ylab("Total number of unique reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = 45000000, linetype = "dashed", color = "red", linewidth= 0.7) +
  geom_text(aes(170, 40000000, label = "Threshold 45M", vjust = -1), color = "red")
  
```



## Duplication rate

```{r}

## graph the combined results
ggscatter(DNA_experiment, x= "Total_DNA", y="Duplication_rate",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 50 ),
          cor.coef.size = 5,
          title = "Impact on duplication rate",
         # facet.by = "Experimenter"
          
          ) +
  ylab("Percentage of duplicate reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5))


## graph the individual results
ggscatter(DNA_experiment, x= "Total_DNA", y="Duplication_rate",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 95 ),
          cor.coef.size = 5,
          title = "Impact on duplication rate",
          facet.by = "Experimenter"
          
          ) +
  ylab("Percentage of duplicate reads") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Number of detected PASS mutations


```{r}

## graph the combined results
ggscatter(DNA_experiment, x= "Total_DNA", y="num_mutations",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 300 ),
          cor.coef.size = 5,
          title = "Impact on PASS mutation detection",
         # facet.by = "Experimenter"
          
          ) +
  ylab("Total number of detected PASS mutations") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5))


## graph the individual results
ggscatter(DNA_experiment, x= "Total_DNA", y="num_mutations",
          color = "black", shape = , size = 3,
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = TRUE,
          cor.coef = TRUE,
          cor.coeff.args = list(method = "pearson", label.sep ="\n", label.x= 90, label.y = 310 ),
          cor.coef.size = 5,
          title = "Impact on PASS mutation detection",
          facet.by = "Experimenter"
          
          ) +
  ylab("Total number of detected PASS mutations") +
  xlab("Total ng of DNA") +
  scale_x_continuous(breaks=c(15, 30, 50, 100, 150, 200)) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Mutations detected in lower amount of DNA samples
What are the mutations detected in 15ng that are not detected in 200 ng?
```{r}
vcf_mutation_ID

vcf_mutation_ID_15ng_vs_200ng <- vcf_mutation_ID %>%
  mutate("15ng" = ifelse((EG_15ng_N03 == TRUE | TS_15ng_N03 == TRUE), TRUE, FALSE)) %>%
  mutate("200ng" = ifelse((EG_200ng_N03 == TRUE | TS_200ng_N03 == TRUE), TRUE, FALSE)) %>%
  select(mutation_ID, "15ng", "200ng") %>%
  filter(`15ng`)  ## removes row that only contain NA (meaning the mutation was detected at 30ng or 50ng)


vcf_mutation_ID_15ng_vs_200ng

vcf_mutation_ID_15ng_ONLY <- vcf_mutation_ID_15ng_vs_200ng %>%
  filter(is.na(`200ng`)) 

vcf_mutation_ID_15ng_ONLY

```




## 15ng vs 200ng
### What FILTER do the mutations not found in 200ng get assigned?
```{r}
#df_AF_total

#vcf_mutation_ID_15ng_ONLY

mutations_ID_GENE_names <- df_AF_total %>%
  select(mutation_ID,GENE) %>%
  unique()

mutations_15ng_ONLY <- left_join(vcf_mutation_ID_15ng_ONLY,df_AF_total) %>%
  filter(Tumor_percentage =="100_%") %>%
  select(mutation_ID,GENE, Total_DNA, FILTER) %>%
  filter(Total_DNA == "200ng") %>%
  unique()

mutations_15ng_ONLY


summarized_mutations_15ng_ONLY <- mutations_15ng_ONLY %>%
  group_by(FILTER) %>%
  summarize(count = n())

summarized_mutations_15ng_ONLY
```


### Do all of these 15ng-only mutations get assigned a SOURCE_ID?
```{r}
mutations_15ng_ONLY_SOURCE_ID <- df_AF_total %>%
  filter(Total_DNA == "200ng" | Total_DNA == "15ng", Tumor_percentage =="100_%", FILTER == "PASS") %>%
  select(mutation_ID,Total_DNA, Experimenter, SOURCE_ID) %>%
  group_by(mutation_ID, Total_DNA) %>%
  summarize(SOURCE_ID_str = paste0(SOURCE_ID, collapse = "_")) %>%
  filter(SOURCE_ID_str != "NA" & SOURCE_ID_str != "NA_NA") %>%
  pivot_wider(names_from = "Total_DNA", values_from = "SOURCE_ID_str", names_prefix = "SOURCE_ID_")

mutations_15ng_ONLY_SOURCE_ID <- mutations_ID_GENE_names %>%
  right_join(vcf_mutation_ID_15ng_ONLY) %>%
  left_join(mutations_15ng_ONLY_SOURCE_ID) %>%
  arrange(SOURCE_ID_15ng)

mutations_15ng_ONLY_SOURCE_ID


```
### Is there a difference in the mean AF value on these mutations?
```{r}
mutations_15ng_ONLY_mean_AF <- df_AF_total %>%
  filter(Total_DNA == "200ng" | Total_DNA == "15ng", Tumor_percentage =="100_%", FILTER == "PASS") %>%
  select(mutation_ID,Total_DNA, Experimenter, AF) %>%
  group_by(mutation_ID, Total_DNA) %>%
  summarize(AF_mean = mean(AF)) %>%
  pivot_wider(names_from = "Total_DNA", values_from = "AF_mean", names_prefix = "mean_AF_") %>%
  arrange(desc(mean_AF_200ng)) %>%
  left_join(mutations_ID_GENE_names) %>%
  relocate(GENE, .after = "mutation_ID")

mutations_15ng_ONLY_mean_AF

mutations_15ng_ONLY_mean_AF <- df_AF_total %>%
  filter(Total_DNA == "200ng" | Total_DNA == "15ng", Tumor_percentage =="100_%", FILTER == "PASS") %>%
  select(mutation_ID,Total_DNA, Experimenter, AF) %>%
  group_by(mutation_ID, Total_DNA) %>%
  summarize(AF_mean = mean(AF)) 

mutations_15ng_ONLY_mean_AF$Total_DNA <- factor(mutations_15ng_ONLY_mean_AF$Total_DNA, levels = c("200ng", "15ng"))


gghistogram(mutations_15ng_ONLY_mean_AF, x = "AF_mean",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 50,
   color = "Total_DNA", fill = "Total_DNA",
   palette = c("#383F51", "#F6E27F")
   )

```

### What are the AF values of the 4 15ng-only mutations that are assigned SOURCE_ID?
```{r}
mutations_15ng_ONLY_SOURCE_ID %>%
  filter(!is.na(SOURCE_ID_15ng)) %>%
  inner_join(mutations_15ng_ONLY_mean_AF)

```



# Project 2: Impact of tumor purity
## FILTER type exploration
### Distribution of main FILTER types

```{r}
# Summarize data by "Sample"
summarized_df <- df_AF_final %>%
  group_by(Total_DNA, Tumor_percentage, Experimenter) %>%
  dplyr::count(FILTER, name = "counts") %>%
  ungroup() %>%
  group_by(Total_DNA, Tumor_percentage,FILTER) %>%
  summarize(mean_counts = mean(counts))

## add category "Other" for low counts
summarized_df <- summarized_df %>%
  mutate(FILTER_new = ifelse(FILTER %in% c("LongHomopolymer", "PASS"), FILTER, "Other" )) %>%
  mutate(FILTER_new = factor(FILTER_new, levels = c("Other", "LongHomopolymer", "PASS"))) ## reorder the values descending

summarized_df

```


```{r}

ggbarplot(summarized_df, x = "Tumor_percentage", y = "mean_counts", fill = "FILTER_new", color = "FILTER_new",
          palette = c("#40798C", "#CFD7C7", "#70A9A1" ),
          facet.by = "Total_DNA") +
    ylab("average counts (n=2)")


```

### What are the PASS mutation of 100% tumor purity that are lost in lower purity samples?
```{r}

## combine replicates and different amounts of starting DNA to see which mutations has been detected in each percentage 
vcf_mutation_ID_100percent_vs_rest <- vcf_mutation_ID %>%
  mutate("100_%" = ifelse((EG_200ng_N03 == TRUE | TS_200ng_N03 == TRUE | EG_15ng_N03 == TRUE | TS_15ng_N03 == TRUE| EG_50ng_N03 == TRUE | TS_50ng_N03 == TRUE | EG_30ng_N03 == TRUE | TS_30ng_N03 == TRUE), TRUE, FALSE)) %>%
  mutate("50_%" = ifelse((`EG_1-2-200ng_N03` == TRUE | `TS_1-2-200ng_N03` == TRUE | `EG_1-2-50ng_N03` == TRUE | `TS_1-2-50ng_N03` == TRUE| `EG_1-2-30ng_N03` == TRUE | `TS_1-2-30ng_N03` == TRUE), TRUE, FALSE)) %>%
  mutate("25_%" = ifelse((`EG_1-4-200ng_N03` == TRUE | `TS_1-4-200ng_N03` == TRUE | `EG_1-4-50ng_N03` == TRUE | `TS_1-4-50ng_N03` == TRUE| `EG_1-4-30ng_N03` == TRUE | `TS_1-4-30ng_N03` == TRUE), TRUE, FALSE))%>%
  select(mutation_ID, "100_%", "50_%", "25_%") %>%
  filter(rowSums(is.na(.)) > 0) %>%  ## removes mutations that have been detected in all 3 percentages
  left_join(mutations_ID_GENE_names, by = "mutation_ID") %>% ## add GENE information
  relocate(GENE, .before = mutation_ID) %>%
  arrange(`100_%`)
  
vcf_mutation_ID_100percent_vs_rest

## mutations are either not detected at all or assigned other filters than PASS


```
```{r}
vcf_mutation_ID_100percent_ONLY <- vcf_mutation_ID_100percent_vs_rest %>%
  filter(`100_%` & is.na(`50_%`)) %>%
  filter(is.na(`25_%`))



mutation_ID_100percent_ONLY <- df_AF_total %>%
  select(mutation_ID,GENE, Tumor_percentage, FILTER) %>%
  filter(Tumor_percentage != "100_%") %>%
  unique() %>%
  pivot_wider(names_from = Tumor_percentage, values_from = FILTER, names_prefix = "FILTER_") %>%
  right_join(vcf_mutation_ID_100percent_ONLY)

mutation_ID_100percent_ONLY



summarized_mutation_ID_100percent_ONLY <- mutation_ID_100percent_ONLY %>%
  pivot_longer(c(`FILTER_50_%`, `FILTER_25_%`),names_to = "FILTER_percentage", values_to = "FILTER" ) %>%
  group_by(FILTER) %>%
  summarize(count = n()) %>%
  mutate(percentage = round((count / sum(count)) * 100, 0)) %>%
  select(-count) %>%
  arrange(desc(percentage))

summarized_mutation_ID_100percent_ONLY$FILTER <- as.character(summarized_mutation_ID_100percent_ONLY$FILTER)

summarized_mutation_ID_100percent_ONLY


```





### Distribution of "Other" FILTER type 
Get the count of AF per each sample
```{r}
## add category "Other" for low counts
summarized_df_other <- summarized_df %>%
  filter(FILTER_new == "Other") %>%
  group_by("Tumor_percentage", "Total_DNA")%>%
  arrange(desc(mean_counts), .by_group = T) %>%    ## reorder the values descending
  mutate(FILTER = as.factor(FILTER))

summarized_df_other
```


```{r}
ggdotchart(summarized_df_other, x = "FILTER", y = "mean_counts",
           color = "Tumor_percentage",                                # Color by groups
           palette = c("#383F51","#00AFBB", "#F6E27F"), # Custom color palette
           sorting = "desc",                       # Sort value in descending order
           add = "segments",                             # Add segments from y = 0 to dots
           rotate = TRUE,                                # Rotate vertically
           group = "Tumor_percentage",                                # Order by groups
           dot.size = 2,                                 # Large dot size
           #label = round(summarized_df_other$mean_counts),                        # Add mpg values as dot labels
           #font.label = list(color = "black", size = 8, vjust = 0.5),               # Adjust label parameters
           ggtheme = theme_pubr(base_size = 8),
           facet.by = "Total_DNA"
           ) + 
  xlab("") +
  ylab("average counts (n=2)")

```




## AF exploration
### Histogram of all detected mutations (PASS and Other)
Lets graph a histogram with the distributions of AF in the samples
```{r}

gghistogram(df_AF_final, x = "AF",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 50,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA")
   )


gghistogram(df_AF_final, x = "AF",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 50,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA", "Experimenter")
   )
```




### histogram of PASS only mutations
Filter for PASS
```{r}
df_AF_final_PASS <- df_AF_final %>%
  filter(df_AF_final$FILTER == "PASS")

percentage_PASS <- round(nrow(df_AF_final_PASS)/nrow(df_AF_final)*100)
print(paste0("Total mutation = ", nrow(df_AF_final), " ; PASS mutations = ", nrow(df_AF_final_PASS), " ; (", percentage_PASS, "%)" ))
```

Plot
```{r}
gghistogram(df_AF_final_PASS, x = "AF",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 50,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA")
   )


gghistogram(df_AF_final_PASS, x = "AF",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 50,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA", "Experimenter")
   )
```

### Statistics on PASS mutations (Violin plots)
```{r}
my_comparisons <- list( c(1, 2), c(1, 3) ) ## pairs to compare

ggviolin(df_AF_final_PASS, x = "Tumor_percentage", y = "AF", fill = "Tumor_percentage",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Total_DNA"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")  # Add significance levels


ggviolin(df_AF_final_PASS, x = "Total_DNA", y = "AF", fill = "Total_DNA",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Tumor_percentage"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")  # Add significance levels

```

### How many mutations are below the 5% AF threshold?

```{r}
df_AF_final_PASS_AF_lower_5_percent <- df_AF_final_PASS %>%
  select(mutation_ID, Tumor_percentage, Total_DNA, Experimenter, AF) %>%
  filter(AF>=0.05) %>%
  group_by( Tumor_percentage, Total_DNA, Experimenter) %>%
  summarize(count = n()) %>%
  group_by(Tumor_percentage, Total_DNA) %>%
  summarize(mean_count =mean(count))



ggbarplot(df_AF_final_PASS_AF_lower_5_percent, x = "Tumor_percentage", y = "mean_count", fill = "Tumor_percentage",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Total_DNA")) +
         ylab("num of mutations with AF<5%")


ggbarplot(df_AF_final_PASS_AF_lower_5_percent, x = "Total_DNA", y = "mean_count", fill = "Total_DNA",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Tumor_percentage"))+
         ylab("num of mutations with AF<5%")


```




## Distubution plot of FQV (Fisher Q Value)
### Histogram of PASS mutations

Lets graph a histogram with the distributions of FPV in the samples
```{r}

gghistogram(df_AF_final_PASS, x = "FQV",
   xscale = "log10",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 60,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA")
   )


gghistogram(df_AF_final_PASS, x = "FQV",
   xscale = "log10",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 60,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Total_DNA", "Experimenter")
   )
```
### Statistical analysis of FQV using violin plots
```{r}
my_comparisons <- list( c(1, 2), c(1, 3) ) ## pairs to compare

ggviolin(df_AF_final_PASS, x = "Tumor_percentage", y = "FQV", fill = "Tumor_percentage",
         yscale = "log10",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Total_DNA"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")  # Add significance levels



ggviolin(df_AF_final_PASS, x = "Total_DNA", y = "FQV", fill = "Total_DNA",
         yscale = "log10",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Tumor_percentage"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")   # Add significance levels





```



## Distubution plot of DP (Depth Coverage)
### Histogram of PASS mutations
Lets graph a histogram with the distributions of AF in the samples
```{r}

gghistogram(df_AF_final_PASS, x = "DP",
   add = "mean", rug = TRUE, add_density = FALSE,
   bins = 60,
   color = "Tumor_percentage", fill = "Tumor_percentage",
   palette = c("#383F51","#00AFBB", "#F6E27F"),
   facet.by=c("Tumor_percentage", "Total_DNA")
   )
```
### Statistical analysis of DP using violin plots
```{r}
my_comparisons <- list( c(1, 2), c(1, 3) ) ## pairs to compare

ggviolin(df_AF_final_PASS, x = "Tumor_percentage", y = "DP", fill = "Tumor_percentage",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Total_DNA"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")  # Add significance levels



ggviolin(df_AF_final_PASS, x = "Total_DNA", y = "DP", fill = "Total_DNA",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         add = "boxplot", add.params = list(fill = "white"),
         facet.by=c("Tumor_percentage"))+
  stat_compare_means(comparisons = my_comparisons, label = "p.signif")   # Add significance levels


```




## What type of mutation changes the most?
```{r}
# Extract mutation info
df_mutation_types <- df_AF_final_PASS %>%
  select(mutation_ID, Total_DNA, Tumor_percentage, Experimenter, REF, ALT) %>%
  unite("mutation_type", REF, ALT, sep = ">") %>%
  group_by(Total_DNA, Tumor_percentage, Experimenter) %>%
  dplyr::count(mutation_type, name = "counts") %>%
  ungroup() %>%
  group_by(Tumor_percentage, mutation_type) %>%
  summarize(mean_counts = mean(counts)) %>%
  filter(mean_counts>4)%>% ## threshold to remove low counts
  arrange(desc(mean_counts))


ggbarplot(df_mutation_types, x = "mutation_type", y = "mean_counts", fill = "Tumor_percentage",
         palette = c("#383F51","#00AFBB", "#F6E27F"),
         drop = FALSE,
         position = position_dodge(0.7),
         #facet.by=c("Tumor_percentage"),
         ylab = "mean counts (replicates, total DNA)",
         xlab = "",
         )+ theme(axis.text.x=element_text(angle = -90, hjust = 0, size = 12))
```









## EXTRA Extrapolation of nucleotides counts (FG and BG) from vcf dataframe

```{r eval=FALSE, include=FALSE}
## extrapolated nucleotide counts for BG
BG_nucleotide_df <- df_AF_final_PASS %>%
  select(mutation_ID, Sample, BG) %>%
  separate(BG, into = c("BG_keys", "BG_values") , sep = ":", remove = FALSE)
  
#unique_key <- "A|C|G|T|D|I"

for (unique_key in unique(BG_nucleotide_df$BG_keys)) { ## loops over unique values of the key (""A|C|G|T|D|I" or "AAAGA|AAGGA|ATT|T"")
  temp_nucleotide_extrapolation_df <- BG_nucleotide_df %>%
    select(mutation_ID, Sample,BG, BG_keys, BG_values) %>%
    filter(BG_keys == unique_key)
  
  string_colnames <- str_split(unique_key, "\\|")[[1]]  ## generates a string with names for the columns in next step ("A", "C" etc)
  
  ## initialization of strings
  col_names <- c()
  index_count <- 0
  
  for (i in string_colnames) {  ## creates a string with colnames of nucelotide and adds a suffix
    index_count <- index_count +1
    col_names[index_count] <- paste0("BG_", i)
  }
  
  temp_nucleotide_extrapolation_df <- temp_nucleotide_extrapolation_df %>% 
    separate(BG_values, into = col_names, sep = "\\|") %>%
    select(-BG, -BG_keys)
  
  
  BG_nucleotide_df <- BG_nucleotide_df%>%
  left_join(temp_nucleotide_extrapolation_df, by = c("mutation_ID", "Sample"), relationship = "many-to-many")

  
}

BG_nucleotide_df <- BG_nucleotide_df%>%
    select(-BG_keys, - BG_values, -BG_AAAG, -BG_AAAGA)

BG_nucleotide_df


## extrapolated nucleotide counts for FG
FG_nucleotide_df <- df_AF_final_PASS %>%
  select(mutation_ID, Sample, FG) %>%
  separate(FG, into = c("FG_keys", "FG_values") , sep = ":", remove = FALSE)
  
#unique_key <- "A|C|G|T|D|I"

for (unique_key in unique(FG_nucleotide_df$FG_keys)) { ## loops over unique values of the key (""A|C|G|T|D|I" or "AAAGA|AAGGA|ATT|T"")
  temp_nucleotide_extrapolation_df <- FG_nucleotide_df %>%
    select(mutation_ID, Sample,FG, FG_keys, FG_values) %>%
    filter(FG_keys == unique_key)
  
  string_colnames <- str_split(unique_key, "\\|")[[1]]  ## generates a string with names for the columns in next step ("A", "C" etc)
  
  ## initialization of strings
  col_names <- c()
  index_count <- 0
  
  for (i in string_colnames) {  ## creates a string with colnames of nucelotide and adds a suffix
    index_count <- index_count +1
    col_names[index_count] <- paste0("FG_", i)
  }
  
  temp_nucleotide_extrapolation_df <- temp_nucleotide_extrapolation_df %>% 
    separate(FG_values, into = col_names, sep = "\\|") %>%
    select(-FG, -FG_keys)
  
  
  FG_nucleotide_df <- FG_nucleotide_df%>%
  left_join(temp_nucleotide_extrapolation_df, by = c("mutation_ID", "Sample"), relationship = "many-to-many")

  
}

FG_nucleotide_df <- FG_nucleotide_df%>%
    select(mutation_ID, Sample, FG_A, FG_C, FG_G, FG_T = `FG_T.x`, FG_D, FG_I)

FG_nucleotide_df




## finally add to the combined table

nucleotide_heatmap_df <- BG_nucleotide_df%>%
    inner_join(FG_nucleotide_df) %>% 
    select(-BG)

nucleotide_heatmap_df


```








































